---
title: "CS5801 Coursework Template Proforma"
author: "student-id"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document: default
  html_document:
    df_print: paged
version: 1
---

# 0. Instructions 

```{r}
library(ggplot2)
library(psych)
library(ggpubr)
library(mgcv)
library(tree)
```


# 1. Organise and clean the data

## 1.1 Subset the data into the specific dataset allocated


```{r}
# Assign your student id into the variable SID, for example:
SID <- 2203898                  # This is an example, replace 2101234 with your actual ID
SIDoffset <- (SID %% 100) + 1    # Your SID mod 100 + 1

load("house-analysis.RDa")
# Now subset the housing data set
# Pick every 100th observation starting from your offset
# Put into your data frame named mydf (you can rename it)
mydf <- house.analysis[seq(from=SIDoffset,to=nrow(house.analysis),by=100),]
```

#Eyeballing the data - Extracting the data to Excel CSV file 
```{r}
write.csv(mydf,"Dataset.csv")

```


## 1.2 Data quality analysis

#Data Quality Checks
#Validity of data  

1)Mandatory: It is mandatory for certain columns like price,mq,floor,n_rooms not to be empty 

2)Missing Data: Checking for implausible , incomplete data values like number of rooms = -1
The number of rooms or bathrooms cannot be -1. It is implausible.

3)Spelling/Numerical Errors : Data observation must not have incorrect notation eg Autonamous instead of Autonomous

4)Data-type: each value has a specific relevant data type eg heating = character string ie autonomous and other
whereas price,mq,n_rooms are numeric data type

5)Range: Minimum and maximum values for numerical variables like price, n_rooms, n_bathrooms, floor, sq

6)Uniqueness: unique variables should be present in the dataset. Checking for duplicated values.

7)Set-membership constraint: Values must come from a set of discrete values/codes 

8) Referential integrity: Constraints based upon multiple variables, e.g., number of floors with respect to total square meter of property eg 14 rooms cannot have 95 sq metre property.

9) Accuracy: accuracy of data with respect to standard values e.g. 1 floor and 2 room flat cannot be 0 sq meter in area. 
 
10)Completeness: comprehensiveness of the data. Entire data set must deliver useful information and must be complete in itself.

11)Consistency: Ensuring variables measured in the same way from different locations etc eg Price in dollars, area in square meters. Same units of measurement ie universal acceptance.
 
12)Traceability: Data must be traceable(sourced).
 
13)Timeliness: Data updates with time eg Price of property changes over time.It depends on inflation and recession.

14)Variance 
almost-zero-variance-features: cases when instance have the same value eg variance between id and property prices. 

15)Outliers: these are abnormal or unusual values that is of significant importance.


 
## 1.3 Data cleaning  

# a) Viewing the data set mydf and column names

```{r}

View(mydf)
names(mydf)

```
 

# b) Examining the first 6 rows and last 6 rows of the data set

```{r}

head(mydf)
tail(mydf)

```

# c) Looking at the structure of the data frame mydf

Inference : 
  - Variables has_terrace, has_alarm, has_air_conditioning, has_parking, is_furnished have data type as "Integers". This needs to be converted to data type "Factor"
  - Heating variable needs to be converted from "character" data type to "factor" data type 
factor conversion using as.factor

```{r}

str(mydf)

```


# Converting given columns to factors
```{r}

mydf$has_terrace<-as.factor(mydf$has_terrace)
mydf$has_alarm<-as.factor(mydf$has_alarm)
mydf$heating<-as.factor(mydf$heating)
mydf$has_air_conditioning<-as.factor(mydf$has_air_conditioning)
mydf$has_parking<- as.factor(mydf$has_parking)
mydf$is_furnished<-as.factor(mydf$is_furnished)
class(mydf$has_alarm)


```


# d) Checking of variables by using table function

```{r}

table(mydf$floor)
table(mydf$n_rooms)
table(mydf$n_bathrooms)
table(mydf$has_terrace)
table(mydf$has_alarm)
table(mydf$heating)
table(mydf$has_air_conditioning)
table(mydf$has_parking)
table(mydf$is_furnished)

```

# d.1)  Running table(mydf$floor) gives
14th and 15th floor which probably is an outlier.
  1   2   3   4   5   6   7   8  14  15 
441 296  98  32  18   7   7   2   1   1 
Since the frequency of both 14th and 15th floor is 1. We can ignore the outlier.

# d.2)  Running table(mydf$n_rooms) gives -1 value :
 -1   2   3   4   5 
  1 155 319 267 161 
This is manual typing error and needs to be changed to 1

```{r}

# Display row from the data frame with incorrect value for n_rooms 
subset(mydf, n_rooms == -1)

#changing -1 to 1
mydf$n_rooms[mydf$n_rooms== -1 ] <- 1
table(mydf$n_rooms)

```

# d.3)  Running table(mydf$heating) gives three outputs : 
autonamous autonomous other 
It is a spelling mistake for autonomous hence we correct the same

```{r}

# Display row from the data frame with incorrect value for Autonomous
subset(mydf, heating == "autonamous")

#changing the name of heating 
mydf$heating[mydf$heating=="autonamous"] <- "autonomous"
table(mydf$heating)

```

# e) Checking for duplicate values

```{r}

duplicated(mydf)
table(duplicated(mydf))

# Duplicated values are absent.

```

# f) Checking for missing values

```{r}

colSums(is.na(mydf))

# Missing values are absent.

```

# g) Checking for numerical summaries

```{r}

summary(mydf)

```

# Summary Insights : Square meter ie mq variable cannot be 0.
Hence we replace 0 with mean value for mq

```{r}

# Display row from the data frame with incorrect value for Square meter
subset(mydf, mq == 0 )

# Replacing 0 with median value of mq
median(mydf$mq)

mydf$mq[mydf$mq== 0] <- 100

summary(mydf)

```


# 2. Exploratory Data Analysis (EDA)

## 2.1 EDA plan

# The Exploratory Data Analysis steps are :
"Price" column is our dependent or response variable. Mostly all graphical insights will be centered towards "Price".
a) Structure of dataframe :
  - Dimensions of data frame
  - Variable names 
  - Data type of each variable
  - Comprehensive understanding of the data frame

b) Numerical Summary of data :
  - Checking Numerical parameters of Descriptive Statistics ie Minimum value, 1st Quantile, Median, Mean, 3rd Quantile, Maximum. This helps to understand abnormal values or any outliers. 

c) Graphical Summaries :
  - Descriptive statistics using visualizations.
  - Histogram for frequency distribution of numerical variables and checking if data is normally distributed.
  - Scatter plots to check correlation between 2 numerical variables.Are they positively, negatively or not correlated.
  - Scatter plots : Slope/ relation => linear, quadratic , exponential  
  - Box plots for categorical and factorial variables with Price.

d) Checking for Normality :
  - Normality test :  shapiro test , histograms
  - Checking if the data falls under normal distribution or is positively or negatively skewed.
  - Using psych library for checking skewness
  - Finding possible reasons for skewness eg outliers

e) Numerically checking for Correlation between variables using cor.test

f) Understanding the research question and formulating appropiate hypothesis.


## 2.2 EDA and summary of results  

# Exploratory Data Analysis
a) Understanding Data frame 

Insights : 
 - Total 903 observations ie rows and 12 columns.
 - Data type of price is numerical and rest other variables are numerical and converted to factors.
 - Factor levels : 0 = absent 1 = present
 
```{r}

str(mydf)

```

b) Numerical summary of Descriptive statistics

Insights :
  - price variable : Mean > Median Value implies positively skewed data, might be indicative of outliers.
  - mq variable : Mean > Median implies positive skewness
  - Numerical Variables like floor, n_rooms, n_bathrooms are discrete values
  - floor : range is greater ie 14 for floor variable. This indicates outliers.
  - Factorial variables : have no known anomalies
  

```{r}

summary(mydf)

# Using psych package 

library(psych) 
describe(mydf)

```

c) Histograms : Numerical variables (checking for frequency distribution, normal distribution)

Insights :
   - price variable : Histogram is slight positive skewed. Nearly normal distribution with maximum frequency of price between 50000-100000.
   - mq variable : Histogram is positively skewed. Outliers present between 400-600 square meters of property.
   - floor variable : edge/outlier present ie 14th floor. Maximum frequency/mode is for 1 floor.
   - n_rooms : Normally distributed. No abnormalities seen. Mode = 3 rooms.
   - n_bathrooms : Descending curve. Frequency of bathrooms = 1 > 2 > 3
   
   
```{r}
# Histogram of price variable
hist(mydf$price, xlab="House Price", ylab="Count", main="Histogram : House Prices", col = "pink") 

# Histogram of mq variable
hist(mydf$mq, xlab="Property Area", ylab="Count", main="Histogram : Property Area", col = "green") 

# Histogram of floor variable
hist(mydf$floor, xlab="Number of Floors", ylab="Count", main="Histogram : Numbers of Floors ", col = "yellow") 

# Histogram of n_rooms variable
hist(mydf$n_rooms, xlab="Number of Rooms", ylab="Count", main="Histogram : Numbers of Rooms ", col = "grey") 

# Histogram of n_bathrooms variable
hist(mydf$n_bathrooms, xlab="Number of Bathrooms", ylab="Count", main="Histogram : Numbers of Bathrooms ", col = "purple") 

```

d) Scatterplots between price variable and other exploratory numerical variables.

Insights :
  - price and mq : Weak positive correlation
  - price and floor : No correlation
  - price and number of rooms : No correlation
  - price and number of bathrooms : No correlation
  

```{r}

library(ggplot2)

# Scatterplot between variables price and mq 
ggplot(data =mydf, aes(x=mq, y=price)) + geom_point() + theme_classic() + ggtitle("Property Price vs Area of property") 

# Scatterplot between variables price and floor
ggplot(data =mydf, aes(x=floor, y=price)) + geom_point() + theme_classic() + ggtitle("Property Price vs Floors") 

# Scatterplot between variables price and n_rooms
ggplot(data =mydf, aes(x=n_rooms, y=price)) + geom_point() + theme_classic() + ggtitle("Property Price vs Number of Rooms") 

# Scatterplot between variables price and n_bathrooms
ggplot(data =mydf, aes(x=n_rooms, y=price)) + geom_point() + theme_classic() + ggtitle("Property Price vs Number of Bathrooms")

```
e) Correlation test : cor.test() between price variable and other exploratory numerical variables.

Insights : Correlation coefficient values are as followings 
  - price and mq : 0.28 implies very weak positive correlation
  - price and floor : 0.03 implies no correlation
  - price and number of rooms : 0.22 implies very weak positive correlation
  - price and number of bathrooms : 0.37 implies weak positive correlation

```{r}

# Correlation between Property Price and Area of Property
cor.test(mydf$price, mydf$mq)

# Correlation between Property Price and Floors
cor.test(mydf$price, mydf$floor)

# Correlation between Property Price and Number of Rooms
cor.test(mydf$price, mydf$n_rooms)

# Correlation between Property Price and Number of Bathrooms
cor.test(mydf$price, mydf$n_bathrooms)

```

f) Boxplots between price and other categorical/factorial exploratory variables

Insights : 
  - price and heating : Almost equal distribution of data in autonomous and other heating type. Autonomous heating has multiple outliers. Median value is approximately same for both.
  - price and has_alarm : House with alarm system have greater price. Greater spread between median and maximum range indicates positive skewness in alarm houses.Houses without alarm system have multiple outliers.
  - price and has_terrace : Median value is approximately same. Non-terraced houses have multiple outliers.
  - price and has_airconditioning :  Median value is approximately same. Non-AC houses have multiple outliers. 
  AC houses have slight positive skew.
  - price and has_parking : Price of houses with parking < Price of houses without parking
                            parking house box plot - positive skew
                            Houses without parking - multiple outliers
  - price and is_furnished : Equal variance in both furnished and not furnished houses. Outliers present in not furnished house. Price range for furnished house < price range for not furnished house
  
```{r}

# Boxplot between price and heating
ggplot(mydf, aes(x=heating, y=price)) + geom_boxplot(fill="steelblue2") + theme_classic() + labs(title="Box Plot of Heating by Property price") 

# Boxplot between price and has_alarm
ggplot(data = mydf, aes(x=has_alarm, y=price)) +geom_boxplot(fill="green") + theme_classic() + ggtitle("Box plot of Price by Alarm system ")

# Boxplot between price and has_terrace
ggplot(data = mydf, aes(x=has_terrace, y=price)) +geom_boxplot(fill="pink") + theme_classic() + ggtitle("Box plot of Price by Terrace")

# Boxplot between price and has_airconditioning 
ggplot(data = mydf, aes(x=has_terrace, y=price)) +geom_boxplot(fill="purple") + theme_classic() + ggtitle("Box plot of Price by Air Conditioning")


# Boxplot between price and has_parking
ggplot(data = mydf, aes(x=has_parking, y=price)) +geom_boxplot(fill="lightgreen") + theme_classic() + ggtitle("Box plot of Price by Parking")

# Boxplot between price and is_furnished
ggplot(data = mydf, aes(x=is_furnished, y=price)) +geom_boxplot(fill="yellow") + theme_classic() + ggtitle("Box plot of Price by Furnished")


```


## 2.3 Additional insights and issues

Additional Insights :
a) Price variable has very weak positive correlation coefficient with other numerical variables - correlation test
b) Price variable has near normal distribution but variables like mq have high positive skewness and outliers. 
c) Floor has extreme outlier ie 15th floor. Since it is not implausible value, we are not deleting the data.
d) Frequency of : 
      - non-terraced flats > terraced flats 
      - flats without alarm system > flats with alarm system
      - flats without air conditioner > flats with air conditioner
      - flats without parking > flats with parking
      - furnished flats > unfurnished flats
e) Autonomous and other heating type have almost equal distribution and median value.
f) There is very less variability of Price explained by exploratory variables graphically. 


# 3. Modelling

## 3.1 Explain your analysis plan

Analysis Plan :

a) Price variable is dependent numerical variable
b) Exploratory variables are both numerical and factorial
c) Multiple regression is an ideal way to analyse the data
d) Before modelling, it is important to explore the data using numerical and graphical summaries
e) Data cleaning should include :
     - Removing invalid, implausible, missing data
f) Data can be corrected by :
      - verifying with the source, imputed, ignored, row deletion, replaced by median  
g) Outliers must not be deleted as they provide meaningful insights
h) Multi-collinearity check - variables using cor function
i) Modelling starts from maximal model and is checked for significance, R-square value, Residual plots
j) Extent of variability explained by the model determines goodness of fit
k) Transformation measures to improve the model
l) AIC (Akaike's Information Criteria) is used to assess model fit

## 3.2 Build a model for property price

Model for property price :
a) Plotting graph to visualize correlation using pairs function
Insights : 
   - Mostly all parameters have zero or positive correlation with each other.
   - Marked variations seen in the data


```{r}

mydf.subset <-subset(mydf, select=c("price", "mq", "floor", "n_rooms", "n_bathrooms")) 
                                  
pairs(mydf.subset)

```

b) Check multi-collinearity between variables
Insights : No multi-collinearity between variables 

```{r}

cor(mydf.subset)

```

c) Maximal model - Multiple Regression 

```{r}

model.1<-lm(mydf$price ~ mydf$mq + mydf$floor + mydf$n_rooms + mydf$n_bathrooms + mydf$has_terrace + mydf$has_alarm + mydf$heating + mydf$has_air_conditioning + mydf$has_parking + mydf$is_furnished)

```


## 3.3 Critique model using relevant diagnostics

a) Model Characteristics : 
  - This model includes all the exploratory variables
  - Coefficients floor and n_bathrooms are most significant followed by has_alarm1
  - R square is less (18.26% )implies only 18.26% of variability is explained by independent variables.
  - F statistic value is less ie 20.28
  - The model is significant
  
b) Diagnostic Tests :
  - Residual vs Fitted graph is slightly curved. 
  - Heteroscedasticity seen - variance unevenly distributed data
  - QQ Plot : S- shaped curve and Outliers present 
  
c) Potential Weakness :
  - There is no significant relationship between price and factorial variables. The presence of factorial variables decreases the R square value.
  - Also, weak correlation coefficient between price and other numerical variables resulted into less fit model
  - Presence of multiple outliers disrupts homogeneous variability of data points
  
  

```{r}

summary(model.1)
plot(model.1)

```



## 3.4 Suggest improvements to your model

Improvements :
a) Transformation - using step function for minimal adequate model
b) Testing minimum adequate model - using summary and diagnostic plots

Insights :
a) Minimal adequate model - property price is significantly dependent on area, number of floors, number of bathrooms, alarm systems
AIC=20556.08
mydf$price ~ mydf$mq + mydf$floor + mydf$n_bathrooms + mydf$has_alarm

b) Model Summary:
    - Model has good acceptance of fit
    - All coefficients have significance
    - R-square is 18%
    - F-statistic is significant ie 50.14

c) Diagnostic Plots:
    - Residual vs Fitted curve has improved. It is linear to horizontal axis
    - Heteroscedasticity seen - variance unevenly distributed
    - QQ Plot : Number of extreme outliers decreased
    - Residual vs Leverage graph : Data points lie within Cook's distance. No significant anomaly seen
    
```{r}

step(model.1)
model.2 <-lm(mydf$price ~ mydf$mq + mydf$floor + mydf$n_bathrooms + mydf$has_alarm)
summary(model.2)
plot(model.2)

```


# 4. Extension work

## 4.1 Model the likelihood of a property being furnished (using the is_furnished variable provided).

Answer 4.1
a) Check data type and frequency of is_furnished variable
Insights : 
   - is_furnished => factorial variable
   - is_furnished has 2 values absent = 0, present = 1
   - Dependent variable is is_furnished = Binary => Logistic Regression
   - Exploratory variables are numerical and factorial
   

```{r}

class(mydf$is_furnished)
table(mydf$is_furnished)

```

b) Logistic Regression using Maximal model without Interactions
Insights : 
  - Intercept and Air Conditioning are significant coefficients
  - AIC - 565.86
  - Diagnostic Plots are not applicable to Logistic regression models
  
```{r}

mydf.logit<-glm(mydf$is_furnished ~ mydf$price + mydf$mq + mydf$floor + mydf$n_rooms + mydf$n_bathrooms + mydf$has_terrace + mydf$has_alarm + mydf$heating + mydf$has_air_conditioning + mydf$has_parking , family=binomial)
summary(mydf.logit)

```

c) Logistic Regression with Interactions 
Insights :
  - Multiple significant coefficients
  - Larger analysis table 

```{r}

mydf.max.logit<-glm(mydf$is_furnished ~ mydf$price * mydf$mq * mydf$floor * mydf$n_rooms * mydf$n_bathrooms * mydf$has_terrace * mydf$has_alarm * mydf$heating * mydf$has_air_conditioning * mydf$has_parking , family=binomial)
summary(mydf.max.logit)

```

d) Logistic Regression using Step Function on Maximal model without Interactions
Insights :
  - Only has_air_conditioning coefficient is significant after performing step function


```{r}

step(mydf.logit)


```


e) Logistic Regression with has_air_conditioning - Minimal adequate model
  
```{r}

mydf.step.logit <- glm(formula = mydf$is_furnished ~ mydf$has_air_conditioning, 
    family = binomial)
summary(mydf.step.logit)

```

f) Calculating Odds Ratio on the Maximal model without Interactions

```{r}

exp(coef(mydf.logit))

```


# References  


CS5701 Week 1 Teaching Materials (2022). Lecture 1:  Exploratory Data Analysis. Available at
https://brightspace.brunel.ac.uk, accessed 20 December 2022.

CS5701 Week 4 Teaching Materials (2022). Lecture 4:  Correlation and Regression. Available at
https://brightspace.brunel.ac.uk, accessed 20 December 2022.

CS5701 Week 6 Teaching Materials (2022). Lecture 6:  Multiple Regression. Available at
https://brightspace.brunel.ac.uk, accessed 22 December 2022.

CS5701 Week 7 Teaching Materials (2022). Lecture 6:  Logistic Regression. Available at
https://brightspace.brunel.ac.uk, accessed 24 December 2022.

CS5702 Week 4 Teaching Materials (2022). Lecture 4:   Exploratory data analysis (EDA). Available at
https://brightspace.brunel.ac.uk, accessed 20 December 2022.

CS5702 Week 5 Teaching Materials (2022). Lecture 5:   Data quality. Available at
https://brightspace.brunel.ac.uk, accessed 23 December 2022.

Crawley, M.J. 2015, Statistics: an introduction using R, Second edn, Wiley, Chichester, West Sussex.

Kabacoff, R.I. 2015, R in Action, Second Edition, Manning Publications






